Relationships in the extremes and their influence on competition and coexistence

### 2021-09-23 paper/Paper.Rnw
some more ediing to intro

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
Furthermore, an additional driving reason to quantify the significance of ATAs for species coexistence is the potential for climate change to alter ATAs of environmental variables implicated in coexistence, as follows. 
The consequence of climate change that has the most potential to influence ATAs in the environment the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 
Temporally extensive extreme events can have important ecological impacts for
competing populations where species growth and the maintenance of biodiversity is dependent of phenologically separated environmental signals. A temporally extensive event can equalize the phenologically separated environmental signals, making those phenological signals more correlated. For example, consider two precipitations signals, one in early spring and one in late summer. If a long flood occurs, it's likely that both signals will be close to equal in that year. If extreme floods occur in more years than not, then the two signal will become more correlated in higher precipitation values, a right-tailed ATA. Similarly, if extreme droughts occur in more years than not, then the signals will be more correlated in the lower values, a left-tailed ATA. The same can be said about spatially separated signals and more spatially extensive extreme events. ATA of environmental signals can then introduce or change ATAs in the fitness or traits of species that depend on the signals and potentially affect their coexistence.
%***DAN:
%1) Topic sentence: 
To make the above ideas more concrete, we review a relevant example from the literature that may be
influenced as previously described. 
We introduce a simplified example of an Arizona desert annual plant community from \cite{Chesson_2012}. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
The community follows the criteria required for temporal storage effects to operate. Different species dominate and have high germination rates depending on rainfall timing. For example, Baileya multiradiata dominates if sufficient rainfall is reached in the early germination season. Eriogonum abertianum dominates is sufficient rainfall is reached in the late germination season. Competition is influenced by resource consumption of seed-producing adult plants. Dormant seeds are unaffected by competiton and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the desert annual community also fluctuate year to year. The result is a relationship between the desert annuals' environmental responses (germination) that is symmetrically associated. However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the desert annual response relationship may become more correlated in the higher values or right-tail associated. Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the desert annual environmental response relationship may become more correlated in the lower values or left tail associated.

To make the above ideas more concrete, we review a relevant example from the literature that may be
influenced as previously described. 
We introduce an example of a two species grass-forb plant community (inspired by \cite{Chesson_2012}).
%2) Make it concrete - name the system, name the organisms, name the seasons. 
To contextualize idea of how ATA in environmental response might manifest and modfiy the outcome of coexistence, The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threhold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficeint rain in the late season. Competition is influenced by resource comsumption of growing juvenile plants. Dormant seeds are unaffected by competiton and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. The result is a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated, and, following our simulated example (Figure \ref{fig:pedagog}), the species persist in coexistence. However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the higher values or right-tail associated. Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb environmental response relationship may become more correlated in the lower values or left tail associated. Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

are more temporally extensive than moderate disturbances, then it is more likely that two species will be influenced by the same events. Both species may become more alike (correlated), an important consideration for competition and coexistence. For example, an extreme flood event that lasts across the growth season of two different species and increases recruitment in both species will cause recruitment of the two species to become more correlated in the upper tail end of the distribution. A subsequent change in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). 
%***DAN: Leave the following for the next para
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??
%Although, to our knowledge, the potential importance of climatic changes in ATAs on coexistence has not been studied,
%the importance for coexistence of other aspects of climate change, including changes in means and variability of environmental variables, 
%have been studied in several systems. 

@@

For example, decreased mean growth season precipitation can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}, and as such, \cite{diPaola_2012} predict the collapse of coexistence between Mediterranean tree species as mean rainfall levels are forecasted to decrease. 

>>>>

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
An additional driving reason to quantify the significance of ATAs for species coexistence is the potential for climate change to alter ATAs of environmental variables implicated in coexistence, as follows. 
Among the various implications of climate change, the most potentially influential to ATAs in the environment is the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 
Temporally extensive extreme events can have important ecological impacts for
competing populations where species growth and the maintenance of biodiversity is dependent on phenologically separated environmental signals. A temporally extensive event can equalize the phenologically separated environmental signals, making them more correlated. For example, consider two precipitations signals, one in early spring and one in late summer, that are usually different. If a long flood occurs, it's likely that both signals will be close to equal in that year. If extreme floods occur in more years than not, then the two signals will become more correlated in higher precipitation values, a right-tailed ATA (figure \ref{fig:noise}C). Similarly, if extreme droughts occur in more years than not, then the signals will be more correlated in the lower values, a left-tailed ATA (figure \ref{fig:noise}A). The same can be said about spatially separated signals and more spatially extensive extreme events. ATA of environmental signals can then introduce or change ATAs in the fitness or traits of species that depend on the signals and potentially affect their coexistence.
%***DAN:
%1) Topic sentence: 
To make the above ideas more concrete, we review a relevant example from the literature that may be
influenced as previously described. 
We introduce a simplified example of an Arizona desert annual plant community from \cite{Chesson_2012}. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
The community follows the criteria required for temporal storage effects to operate. Different species dominate and have high germination rates depending on rainfall timing. For example, Baileya multiradiata dominates if sufficient rainfall is reached in the early germination season. Eriogonum abertianum dominates if sufficient rainfall is reached in the late germination season. Competition is influenced by resource consumption of seed-producing adult plants. Dormant seeds are unaffected by competition and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the desert annual community also fluctuate year to year. The result is a relationship between the desert annuals' environmental responses (germination) that is symmetrically associated...

To make the above ideas more concrete, we review a relevant example from the literature that may be
influenced as previously described. 
We introduce a hypothetical example of the commonly studied two species grass-forb plant community \citep{Chesson_2012, Hallett_2019}.
The community follows the criteria required for temporal storage effects to operate. The forb dominates and has high fecundity when a sufficient threshold of rain is reached in the early season. The grass dominates and has high fecundity when there is sufficient rain in the late season. Competition is influenced by resource consumption of growing juvenile plants. Dormant seeds are unaffected by competition and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the grass-forb community also fluctuate year to year. The result is a relationship between the grass's and forb's environmental responses (fecundity) that is symmetrically associated, and the species persist in coexistence...

However, if environmental responses become asymmetrically associated by, say, extreme rainfall events, coexistence can be altered. If the community is disturbed by frequent and long floods with some years of normalcy, the grass-forb/desert annual 
environmental response relationship may become more correlated in the higher values, a right-tailed ATA (figure \ref{fig:noise}C). Conversely, if the community is disturbed by frequent and long drought with some years of normalcy, the grass-forb/desert annual environmental response relationship may become more correlated in the lower values, a left-tailed ATA (figure \ref{fig:noise}A). 
The subsequent potential changes in competitive ability and coexistence is not just attributed to the extreme event alone, but from the resulting extreme relationship between species (ATA). 
%Again, we see the possible consequences of ATA in Figure \ref{fig:pedagog}, i.e., less or no coexistence.

%***DAN: Leave the following for the next para
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

%***JAS: I am trying to relay the message that climate change and specifically extreme events have consequences for coexistence and competition
% Maybe a better way to start this paragraph is:
% EC covariance, and thus the storage effect, may be altered by changes in the environment. 
% need an indication that we are more interested in increased variance and extreme events because we are looking we are looking for changes in a co-varying relationship??

@@

Decreased mean growth season precipitation, for instance, can decrease the competitive ability of dominant forest tree species \citep{Clark_2011}, and coexistence is predicted to collapse between Mediterranean tree species in accordance to forecasted mean rainfall level decrease \cite{diPaola_2012}. 
